---
title: "Overview"
---

<img src = "images/nceas_logo_wide.png" align = "right" width = "50%" />

### Welcome!

This workshop provides an overview of many of the packages included in the `tidyverse` suite of packages for the R programming language. The `tidyverse` is a veritable universe of tools though that no single workshop could hope to cover so **we are focusing here on an introductory approach that focuses primarily on some fundamentals to tidying data in R**. We are always happy to improve workshop content so please don't hesitate to [post an Issue](https://github.com/NCEAS/scicomp-workshop-tidyverse/issues) on our GitHub repository if you see clear areas for improvement!

<img src = "images/hex_tidyverse.png" align = "right" width = "17%" />

To maximize the value of this workshop to you, **we recommend you take the following steps before the day of the workshop**. If anything is unclear, feel free to reach out to us; our contact information can be found in the "Content Creators" tab.

## Programs to Install

### R & RStudio

**Install [R](https://www.r-project.org/) and its more convenient (in our opinion) user-interface: [RStudio](https://www.rstudio.com/products/rstudio/download/)**.

If you already have R, check that you have at least version 4.0.0 by running the following code:

```{r, eval = F}
version$version.string
```

If your version starts with a 3 (e.g., the above code returns "R version 3..."), please update R to make sure all packages behave as expected.

### R Packages

**Install the `tidyverse` and `palmerpenguins` R packages** using the following code:

```{r, eval = F}
install.packages("tidyverse", "palmerpenguins")
library(tidyverse)
library(palmerpenguins)
```

**Please run the above code even if you already have these packages** to update these packages and ensure that your code aligns with the examples and challenges introduced during the workshop.

## Websites to Visit

### Supplemental Material

While not technically necessary to attend the workshop, if you'd like you can see the content that created the workshop website you are viewing by visiting our [GitHub repository here](https://github.com/NCEAS/scicomp-workshop-tidyverse).
