---
title: "Overview"
---

<img src = "images/nceas_logo_wide.png" align = "right" width = "50%" />

### Welcome!

A short blurb on the topic of the workshop should be covered here. Strive to be welcoming and nonthreatening as this page includes our 'before you arrive' instructions and thus precedes workshop participants actually meeting us in person (or via Zoom).


We are always happy to improve workshop content so please don't hesitate to [post an Issue](https://github.com/NCEAS/scicomp-workshop-template/issues) on our GitHub repository if you see clear areas for improvement!

To maximize the value of this workshop to you, **we recommend you take the following steps before the day of the workshop**. If anything is unclear, feel free to reach out to us; our contact information can be found in the "Content Creators" tab.

## Programs to Install

### R & RStudio

**Install [R](https://www.r-project.org/) and its more convenient (in our opinion) user-interface: [RStudio](https://www.rstudio.com/products/rstudio/download/)**.

If you already have R, check that you have at least version 4.0.0 by running the following code:

```{r, eval = F}
version$version.string
```

If your version starts with a 3 (e.g., the above code returns "R version 3..."), please update R to make sure all packages behave as expected.

### R Packages

**Install the `package A` and `package B` R packages** using the following code:

```{r, eval = F}
install.packages("tidyverse", "palmerpenguins")
library(tidyverse)
library(palmerpenguins)
```

**Please run the above code even if you already have these packages** to update these packages and ensure that your code aligns with the examples and challenges introduced during the workshop.

## Websites to Visit

### Supplemental Material

While not technically necessary to attend the workshop, if you'd like you can see the content that created the workshop website you are viewing by visiting our [GitHub repository here](https://github.com/NCEAS/scicomp-workshop-template).
