---
title: "Summarize"
---

```{r, echo = F, message = F}
library(tidyverse); library(palmerpenguins)
```

## Pipe Operator (`%>%`)

<img src = "images/nceas_logo_square.png" align = "right" width = "15%" />

Before diving into the `tidyverse` functions that allow for summarization and group-wise operations, let's talk about the pipe operator (`%>%`). The pipe is from the `magrittr` package and allows chaining together multiple functions without needing to create separate objects at each step as you would have to without the pipe.

### Example: Pipe

As in the other chapters, let's use the "penguins" data object found in the `palmerpenguins` package. Let's say we want to keep only specimens that have a measurement for both bill length and bill depth and then remove the flipper and body mass columns.

Without the pipe--but still using other `tidyverse` functions--we could go about this like this:
```{r, message = F}
# Filter out the NAs
penguins_v2 <- dplyr::filter(.data = penguins,
                              !is.na(bill_length_mm) & !is.na(bill_depth_mm))

# Now strip away the columns we don't want
penguins_v3 <- dplyr::select(.data = penguins_v2, 
                             -flipper_length_mm, -body_mass_g)

# And we can look at our final product with `base::head`
head(penguins_v3)
```

Using the pipe though we can simplify this code dramatically! Note that each of the following lines must end with the `%>%` so that R knows there are more lines to consider.
```{r, message = F}
# We begin with the name of the data object
penguins %>%
  # Then we can filter the data
  dplyr::filter(!is.na(bill_length_mm) & !is.na(bill_depth_mm)) %>%
  # And strip away the columns we don't want
  dplyr::select(-flipper_length_mm, -body_mass_g) %>%
  # And we can even include the `head` argument to see our progress
  head()
```

Note that using the pipe allows each line to inherit the data provided by the previous line.

### Challenge: Pipe

Using pipes, filter the data to only include male penguins, keep only the columns for species, island, and body mass, and drop any rows with NA for body mass.

### Aside: Fun History of Why `%>%` is a "Pipe"

The Belgian painter René Magritte famously created a painting titled "[The Treachery of Images](https://collections.lacma.org/node/239578)" featuring a depiction of a smoking pipe above the words "*Cest ci n'est pas une pipe*" (French for "This is not a pipe"). Magritte's point was about how the depiction of a thing is not equal to thing itself. The `magrittr` package takes its name from the painter because it also includes a pipe that functions slightly differently from a command line pipe and uses different characters. Just like Magritte's pipe, `%>%` both is and isn't a pipe!

## Group-Wise Summarizing

Now that we've covered the `%>%` operator we can use it to do group-wise summarization! Technically this summarization does not *require* the pipe but it does inherently have two steps and thus benefits from using the pipe to chain together those technically separate instructions.

To summarize by groups we first define our groups using `dplyr`'s `group_by` function and then summarize using `summarize` (also from `dplyr`). `summarize` does require you to specify what calculations you want to perform within your groups though it uses similar syntax to `dplyr`'s `mutate` function.

Despite the similarity in syntax between `summarize` and `mutate` there are a few crucial differences:

- `summarize` returns only a single row per group while `mutate` returns as many rows as are in the original dataframe
- `summarize` will automatically remove any columns that aren't either (1) included in `group_by` or (2) created by `summarize`. `mutate` cannot remove columns so it only creates whatever you tell it to.

### Example: Summarize

Let's say we want to find some summary values for body mass of each species of penguin on each island. To accomplish this we can do the following:

```{r, message = F}
# Begin with the data and a pipe
penguins %>%
  # Group by the desired column names
  dplyr::group_by(species, island) %>%
  # And summarize in the way we desire
  dplyr::summarise(
    # Get average body mass
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),
    # Get the standard deviation
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    # Count the number of individual penguins of each species at each island
    n_mass = dplyr::n(),
    # Calculate standard error from SD and count
    se_mass = sd_mass / n_mass )
  
```

You can see we also invoked the `n` function from `dplyr` to return the size of each group. This function reads any groups created by `group_by` and returns the count of rows in the dataframe for each group level.

### Challenge: Summarizing

Using what we've covered so far, find the number of penguins sampled in each year (regardless of any other grouping variable).

## Grouping Cautionary Note

`group_by` can be extremely useful in summarizing a dataframe or creating a new column without losing rows but you need to be careful. Objects created with `group_by` "remember" their groups until you change the groups or use the function `ungroup` from `dplyr`.

Look at how the output of a grouped data object tells you the number of groups in the output (see beneath this code chunk).

```{r, message = F}
penguins %>%
  dplyr::group_by(species, island) %>%
  dplyr::summarize(penguins_count = dplyr::n())
```

This means that all future uses of that pipe will continue to use the grouping established to create the "penguins_count" column. We can stop this by doing the same pipe, but adding `ungroup` after we're done using the grouping established by `group_by`.

```{r, message = F}
penguins %>%
  dplyr::group_by(species, island) %>%
  dplyr::summarize(penguins_count = dplyr::n()) %>%
  dplyr::ungroup()
```

See? We calculated with our desired groups but then dropped the grouping structure once we were finished with them. Note also that if you use `group_by` and do some calculation then re-group by something else by using `group_by` *again*, the second use of `group_by` **will not be affected** by the first. This means that you only need one `ungroup` per pipe.
